---
title: "Blog Post 2 Code"
author: "Vivian Nguyen"
date: '2022-09-17'
output: pdf_document
---

```{r setup, include=FALSE}

## Set up

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(car)
library(readr)
library(stargazer)
library(dplyr)

popvote_df <- read_csv('house_popvote_seats.csv') 
GDPquarterly_df <- read_csv('GDP_quarterly.csv')
RDIquarterly_df <- read_csv('RDI_quarterly.csv')
UEquarterly_df <- read_csv('unemployment_national_quarterly_final.csv')
# UE <- read_csv('unemployment_state_monthly.csv')

```

```{r, include = FALSE, message=FALSE}

## Merge predictor and outcome data, filter for just the last Q of the election cycle

results_GDP <- left_join(GDPquarterly_df, popvote_df, by = 'year') %>% 
  drop_na() %>% 
    select('year', 'winner_party', 'H_incumbent_party', 'H_incumbent_party_majorvote_pct', 'quarter_cycle', 'GDP_growth_pct', 'GDP_growth_qt') %>%
    filter(quarter_cycle == 8)  

results_RDI <- left_join(RDIquarterly_df, popvote_df, by = 'year') %>% 
  drop_na() %>% 
    select('year', 'winner_party', 'H_incumbent_party', 'H_incumbent_party_majorvote_pct', 'quarter_cycle', 'DSPIC_change_pct', 'DSPIC_change_qt') %>%
    filter(quarter_cycle == 8) 

UEquarterly_df$UNRATE_change <- 0 

for (index in 1:nrow(UEquarterly_df)){
  if (index != 1)
    UEquarterly_df$UNRATE_change[index] <- UEquarterly_df$UNRATE[index] - UEquarterly_df$UNRATE[index - 1]
}

results_UE <- left_join(UEquarterly_df, popvote_df, by = 'year') %>% 
  drop_na()  %>% 
    select('year', 'winner_party', 'H_incumbent_party', 'H_incumbent_party_majorvote_pct', 'quarter_cycle', 'UNRATE', 
           'unrate_qt_total', 'UNRATE_change') %>%
    filter(quarter_cycle == 4) 

```

```{r, echo=FALSE, eval=TRUE, fig.width=5, fig.height=3, out.width = '60%', fig.align='center'}

## Scatter plots

# GDP vs. House Incumbent Vote Share
results_GDP %>%
  ggplot(aes(x=GDP_growth_pct, y=H_incumbent_party_majorvote_pct,
             label=year)) + 
    geom_text() +
    geom_hline(yintercept=50, lty=2) +
    geom_vline(xintercept=0.00, lty=2) + # median
    xlab("Q8-Q7 GDP growth") +
    ylab("Incumbent party PV") +
    theme_bw() +
    theme(axis.text = element_text(size = 10))

# RDI vs. House Incumbent Vote Share
results_RDI %>%
  ggplot(aes(x=DSPIC_change_pct, y=H_incumbent_party_majorvote_pct,
             label=year)) + 
    geom_text() +
    geom_hline(yintercept=50, lty=2) +
    geom_vline(xintercept=0.00, lty=2) + # median
    xlab("Q8-Q7 DSPIC growth") +
    ylab("Incumbent party PV") +
    theme_bw() +
    theme(axis.text = element_text(size = 10))

# UE vs. House Incumbent Vote Share
results_UE %>%
  ggplot(aes(x=UNRATE, y=H_incumbent_party_majorvote_pct,
             label=year)) + 
    geom_text() +
    geom_hline(yintercept=50, lty=2) +
    geom_vline(xintercept=0.00, lty=2) + # median
    xlab("Q8-Q7 UE rate") +
    ylab("Incumbent party PV") +
    theme_bw() +
    theme(axis.text = element_text(size = 10))

```

## Bivariate correlation of GDP /  and PV

```{r, eval=FALSE}

## Finding correlations

# cpi <- read_csv("CPI_monthly.csv")

cor(results_GDP$GDP_growth_pct, results_GDP$H_incumbent_party_majorvote_pct)
cor(results_RDI$DSPIC_change_pct, results_RDI$H_incumbent_party_majorvote_pct)
cor(results_UE$UNRATE, results_UE$H_incumbent_party_majorvote_pct)

```

## Economy and PV: Fitting a model (STEP 1 \& 2) {.build}

```{r}

newdf <- left_join(results_UE, results_GDP, by = 'year')
newdf <-left_join (newdf, results_RDI, by = 'year') %>%
  select('year', 'winner_party.x', 'H_incumbent_party.x', 'H_incumbent_party_majorvote_pct.x', 'quarter_cycle.x', 'UNRATE', 'GDP_growth_pct', 'DSPIC_change_pct', 'unrate_qt_total', 'DSPIC_change_qt', 'GDP_growth_qt', 'UNRATE_change') %>%
  rename('winner_party' = 'winner_party.x', 'H_incumbent_party_majorvote_pct' = 'H_incumbent_party_majorvote_pct.x', 'quarter_cycle' = 'quarter_cycle.x', 'H_incumbent_party' = 'H_incumbent_party.x')

newdf
write.csv(newdf, "house_results_econ_vars.csv", row.names = TRUE)

```


```{r, echo=TRUE}

lm_GDP <- lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct, data = newdf)
lm_RDI <- lm(H_incumbent_party_majorvote_pct ~ DSPIC_change_pct, data = newdf)
lm_UE <- lm(H_incumbent_party_majorvote_pct ~ UNRATE, data = newdf)
lm_model <- lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct + DSPIC_change_pct + GDP_growth_qt + UNRATE, data = newdf)
stargazer(lm_GDP, lm_RDI, lm_UE, lm_model, title = "Results", type = "text")

```


## Economy and PV: Fitting a model (STEP 1 \& 2)


```{r, echo=FALSE, fig.width = 8, fig.height=6, out.width = '100%', fig.align='center'}
# GDP plot
newdf %>%
  ggplot(aes(x=GDP_growth_pct, y=H_incumbent_party_majorvote_pct,
             label=year)) + 
    geom_text(size = 8) +
    geom_smooth(method="lm", formula = y ~ x) +
    geom_hline(yintercept=50, lty=2) +
    geom_vline(xintercept=0.01, lty=2) + # median
    xlab("Q8-Q7 GDP growth (X)") +
    ylab("Incumbent party PV (Y)") +
    theme_bw() +
    ggtitle("Y = 52.42 - 0.9118  * X") + 
    theme(axis.text = element_text(size = 20),
          axis.title = element_text(size = 24),
          plot.title = element_text(size = 32))

print(summary(lm_GDP)$coefficients)
print(summary(lm_RDI)$coefficients)
print(summary(lm_UE)$coefficients)


# RDI plot
newdf %>%
  ggplot(aes(x = DSPIC_change_pct, y = H_incumbent_party_majorvote_pct,
             label=year)) + 
    geom_text(size = 8) +
    geom_smooth(method="lm", formula = y ~ x) +
    geom_hline(yintercept=50, lty=2) +
    geom_vline(xintercept=0.01, lty=2) + # median
    xlab("Q8-Q7 RDI growth (X)") +
    ylab("Incumbent party PV (Y)") +
    theme_bw() +
    ggtitle("Y = 52.58 - 0.7964  * X") + 
    theme(axis.text = element_text(size = 20),
          axis.title = element_text(size = 24),
          plot.title = element_text(size = 32))

# UE plot
newdf %>%
  ggplot(aes(x = UNRATE, y = H_incumbent_party_majorvote_pct,
             label=year)) + 
    geom_text(size = 8) +
    geom_smooth(method="lm", formula = y ~ x) +
    geom_hline(yintercept=50, lty=2) +
    geom_vline(xintercept=0.01, lty=2) + # median
    xlab("Q8-Q7 UE (X)") +
    ylab("Incumbent party PV (Y)") +
    theme_bw() +
    ggtitle("Y = 47.48 + 0.7466082  * X") + 
    theme(axis.text = element_text(size = 20),
          axis.title = element_text(size = 24),
          plot.title = element_text(size = 32))

```


# How to evaluate your model (STEP 3)

## Model Fit: $R^2$

```{r}

summary(lm_GDP)$r.squared
summary(lm_RDI)$r.squared
summary(lm_UE)$r.squared
summary(lm_model)$r.squared

```

## Model Fit: in-sample error and MSE


```{r, fig.width=3, fig.height=2, out.width = '30%', fig.align='center'}
# UE
plot(newdf$year, newdf$H_incumbent_party_majorvote_pct, 
     type="l",
     main="true Y (line), predicted Y (dot) for each year")
points(newdf$year, predict(lm_UE, newdf))

# My model
plot(newdf$year, newdf$H_incumbent_party_majorvote_pct, 
     type="l",
     main="true Y (line), predicted Y (dot) for each year")
points(newdf$year, predict(lm_model, newdf))

```

## Model Fit: in-sample error and MSE

We can plot the in-sample error via \textbf{residuals}, which capture the difference between each observed value ($y_i$) and predicted value ($\widehat{y}_i = \hat{A} + \widehat{B}x_i$):

```{r, fig.width=6, fig.height=4, out.width = '70%', fig.align='center'}
# UE
hist(lm_UE$model$H_incumbent_party_majorvote_pct - lm_UE$fitted.values, 
     main="histogram of true Y - predicted Y")
# My model
hist(lm_model$model$H_incumbent_party_majorvote_pct - lm_model$fitted.values,
     main="histogram of true Y - predicted Y")
```

## Model Fit: in-sample error and MSE

We can summarise the error a single number, such as the \textbf{mean-squared error (MSE)}:

```{r}
# GDP
mse_UE <- mean((lm_UE$model$H_incumbent_party_majorvote_pct - lm_UE$fitted.values)^2)
mse_GDP <- mean((lm_GDP$model$H_incumbent_party_majorvote_pct - lm_GDP$fitted.values)^2)
mse_RDI <- mean((lm_RDI$model$H_incumbent_party_majorvote_pct - lm_RDI$fitted.values)^2)
mse_model <- mean((lm_model$model$H_incumbent_party_majorvote_pct - lm_model$fitted.values)^2)

cat("UE:", sqrt(mse_UE))
cat("\nGDP:", sqrt(mse_GDP))
cat("\nRDI:", sqrt(mse_RDI))
cat("\nMy model:", sqrt(mse_model))

```
\textcolor{red}{This is hard to interpret on its own, more useful in comparison with other models.}

## Model Testing
\pause
Checking in-sample model predictions is a good baseline evaluation, but it feels a bit like "cheating" \pause $\rightsquigarrow$ can we take away the model's "answer key"? \pause

<!--the best test of a model's predictive score -- simulate out-of-sample 
    prediction and check performance against ground-truth; in-sample is sort of 
    easy because it's double dipping the data, this is a harder test!-->

We can simulate \textbf{out-of-sample prediction} (also called out-of-sample testing) by withholding some observation, e.g. $X_{2018}$, before fitting: \pause

```{r}
#My model
outsamp_mod1 <- lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct,
                    newdf[newdf$year != 2018,])
outsamp_pred <- predict(outsamp_mod1,
                        newdf[newdf$year == 2018,])
outsamp_true <- newdf$H_incumbent_party_majorvote_pct[newdf$year == 2018]
```
\pause and see how well the model predicts the true $Y_{2018}$ for the held-out observation $X_{2018}$:
```{r}
outsamp_pred - outsamp_true
```
Leaving a single observation out and testing the model against the ground truth is, \pause you guessed it, called \textbf{leave-one-out validation}.

## Model Testing

\textbf{Cross-validation}: withhold a *random subset* of the sample, fit model on rest of sample, and evaluate predictive performance on the held-out observations. \pause

<!-- Me: predicts 2016 within 2 points..but ok 2016 was weird -->


```{r}
# GDP
years_outsamp <- sample(newdf$year, 8)
mod <- lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct,
          newdf[!(newdf$year %in% years_outsamp),])
outsamp_pred <- predict(mod,
                newdata = newdf[newdf$year %in%
                                 years_outsamp,])
```
\pause
```{r}
mean(outsamp_pred - newdf$H_incumbent_party_majorvote_pct[newdf$year %in% years_outsamp])
```
\pause \textcolor{red}{But we don't want to do this just once.}

## Model Testing

Cross-validation involves repeatedly evaluating performance against many randomly held-out "out-of-sample" datasets:

<!-- we can do this many times and see how out-of-sample performance usually is -->
```{r, eval=FALSE}
years_outsamp <- sample(newdf$year, 8)
outsamp_mod <- lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct, newdf[!(newdf$year %in% years_outsamp),])
outsamp_pred <- predict(outsamp_mod, newdata = newdf[newdf$year %in% years_outsamp,])
outsamp_true <- newdf$H_incumbent_party_majorvote_pct[newdf$year %in% years_outsamp]
mean(outsamp_pred - outsamp_true)
```

## Model Testing

Cross-validation involves repeatedly evaluating performance against many randomly held-out "out-of-sample" datasets:

```{r}
outsamp_errors <- sapply(1:1000, function(i){
    years_outsamp <- sample(newdf$year, 8)
  outsamp_mod <- lm(H_incumbent_party_majorvote_pct ~
                      GDP_growth_pct,
                  newdf[!(newdf$year %in% years_outsamp),])
  outsamp_pred <- predict(outsamp_mod,
                newdata = newdf[newdf$year %in% years_outsamp,])
  outsamp_true <- newdf$H_incumbent_party_majorvote_pct[newdf$year
                                                       %in% years_outsamp]
  mean(outsamp_pred - outsamp_true)
})
```

##  Model Testing

```{r, echo=FALSE, message=FALSE, fig.width=6, fig.height=3, out.width = '70%', fig.align='center'}
hist(outsamp_errors,
     xlab = "",
     main = "mean out-of-sample residual\n(1000 runs of cross-validation)")
```


```{r}
mean(abs(outsamp_errors))
```

## Economy and PV: Out-of-sample prediction (STEP 4 \& 5)

```{r, echo=TRUE, eval=FALSE}

GDP_new <- GDPquarterly_df %>%
    subset(year == 2020 & quarter_cycle == 8) %>%
    select(GDP_growth_pct)
UNRATE_new <- UEquarterly_df %>%
    subset(year == 2020 & quarter_cycle == 4) %>%
    select(UNRATE)
DSPIC_new <- RDIquarterly_df %>%
    subset(year == 2020 & quarter_cycle == 8) %>%
    select(DSPIC_change_pct)
GDP_qt_new <- GDPquarterly_df %>%
    subset(year == 2020 & quarter_cycle == 8) %>%
    select(GDP_growth_qt)

new_data <- data.frame(GDP_new, DSPIC_new, GDP_qt_new, UNRATE_new)

predict(lm_model, new_data, interval="confidence")

#Interpretation: This means that, according to our model, a year with these traits, on average, has a (stopping distance ranging between 51.83 and 62.44 ft)46, 54) H incumbent vote share

```


